---
title: "<br>Writing R Packages"
subtitle: "<br>_Enhance and Advance_"
author: "Eric Dunford "
date: "<br>_McCourt School of Public Policy_<br>Georgetown University"
output:
  xaringan::moon_reader:
    css: [default]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = T,message=F,error=F,warning = F)
```

<br><br><br><br><br><br><br>

.center[
### All slides and materials can be found at 
### [http://ericdunford.com/writing_rpackages](http://ericdunford.com/writing_rpackages)
]

---

### Today will be a brief introduction. For a deeper dive, read...

### [R Packages: _Organize, Test, Document, & Share Your Code_](https://r-pkgs.org/index.html) by Hadley Wickham and Jenny Bryan.

.center[

<img src = "figures/r_packages_cover.png" width=40%, height=40%>
]

---

## When to write an R Package

<br>

### (1) Repeat code _more than once_ &rarr; write a function.

--

### (2) Use functions more than once in a project &rarr; write a function script.

--

### (3) Use functions across projects and tasks &rarr; write a package.

---

## What is an R package

#### (1) Set of **Functions** (and logic)

--

#### (2) **Documentation**

> Clear guidance on how to use your functions

--

#### (3) **Dependencies** 

> Functionality from other packages that you're drawing from;

--

#### (4) Example **Data** 

> Documentation require examples, and examples sometimes require unique data. 

--

#### (5) **Vignettes**/Walkthrough

> Show us how to use your package; outline the package workflow. 

---

### R packges that make writing R packages incredibly easy


### [`roxygen2`](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html)

> Streamlines creating documentation.

.center[

<img src = "figures/roxygen2_logo.png" width=40%, height=40%>
]

---

### R packges that make writing R packages incredibly easy


### [`usethis`](https://usethis.r-lib.org/)

> Automates repetitive tasks during development. 

.center[

<img src = "figures/usethis_logo.png" width=40%, height=40%>
]

---

### R packges that make writing R packages incredibly easy


### [`testthat`](https://testthat.r-lib.org/)

> Writing & running unit tests. 

.center[

<img src = "figures/testthat_logo.png" width=40%, height=40%>
]


---

class: inverse, center, middle

# 10 "Easy" Steps to R Package Development

---

### (1) Have a reason to build a package

- Functions you use over and over again that you want to easily access
- Standardize workflow and processes across your team by building a dedicated library
- Have a novel idea for a method/process/workflow that doesn't exist in R but should

--

### (2) Come up with a name for your package

- Contain only ASCII letters, numbers, and '.'
- Have at least two characters
- Start with a letter
- Not end with '.'
- `available::available("my_package_name", browse = FALSE)`

---

### (3) Build the package infrastructure

```r
# where your package should live
path_to_your_project <- "path/to/a/folder"

# initialize everything you'd need to get started
usethis::create_package(path_to_your_project)

# license your package (so no one can steal 
# & profit from your package)
usethis::use_mit_license("Your Name") 

# Setup unit tests (more on this later)
usethis::use_test("tests")

# Create a readme (to help generate a nice github vignette)
usethis::use_readme_rmd()

# Initialize your git repository
usethis::use_git(message="initial commit of my package")
```

---

### (4) Drop you functions into the `R/` folder. 

> Create a file &rarr; think carefully about naming conventions 

--

### (5) Add Roxygen2 skeletons to each function and fill in.

> shift + alt + cmd + r (when cursor is on the function on a mac in RStudio) OR click on the "wand" icon and select "Insert Roxygen Skeleton"

```r
#' Title
#'
#' @param arg1 
#' @param arg2 
#'
#' @return
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  return("hello")
}
```
---

### (6) Make dependencies explicit

- ` usethis::use_package("dplyr")` &rarr; imports ALL functions from that package.
  - different `type=` are allowed, user "imports".
- Explicitly call to the packages methods in your code using it's namespace reference, e.g. `dplyr::select()`
- Use `usethis::` to add common package dependencies, such as the pipe. `usethis::use_pipe()`

--

### (7) Fill in the `DESCRIPTION.R`

- Add a one sentence description (all in title case) of what your package does. `Title:`
- Add a one paragraph description of what your package does. `Description:`
- Fill in the author(s) information `Authors@R:`
- Down the line, you'll want to iterate on the version number. `Description:`

---

### (8) Design unit tests and run them

- Why? &rarr; more on this in a moment
- To run tests:
  - `devtools::test()`
  - shift + cmd + T (on a mac in RStudio)

--

### (9) Build documentation

- Why? 
  - "Building" the documentation translates all the information in your Roxygen2 skeleton to a special `.Rd` (R Documentation) file housed in the `man/` folder. 
  - Documentation is required for every exported function
  - This is what pops up in the "help" pane when you run `?function_name()`
- How?
  - `devtools::document(roclets = c('rd', 'collate', 'namespace'))`
  - shift + cmd + D (on a mac in RStudio) 
  - "Build" Tab in R studio &rarr; click on "more" drop down  &rarr; click on "document"

---

### (10) Build the package

- Why?
  - "Building" installs your package on your machine (so you can use it like any other package).
- How?
  - commandline: `R CMD INSTALL --no-multiarch --with-keep.source YourPackageName`
  - shift + cmd + B (on a mac in RStudio)
  - "Build" Tab in R studio &rarr; click on "Install and Restart".
- When you're ready to deploy your package to CRAN:
  - "Build" Tab in R studio &rarr; click on "more" drop down  &rarr; click on "Build Source Package"
  - This generates a `.tar.gz` file. This is the file you submit to CRAN. 

---

class: inverse, center, middle

# Regarding Roxygen Skeletons

---

### Let's break this down

```{r, eval=F,highlight=T}
#' Title
#'
#' @param arg1 
#' @param arg2
#'
#' @return
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```


---

### Let's break this down

```{r, eval=F,highlight=T}
#' Title
#'
#' @param arg1 
#' @param arg2
#'
#' @return
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---


### The function name is at the top with a description of what the function does beneath it.


```{r, eval=F,highlight=T}
#' my_function() #<<
#'
#' This function pastes together two inputs in a string #<<
#'
#' @param arg1 
#' @param arg2
#'
#' @return
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---


### `@param` describes each function input (what it does, valid inputs, default values)

```{r, eval=F,highlight=T}
#' my_function() 
#'
#' This function pastes together two inputs in a string 
#'
#' @param arg1 a string to be pasted; default is NA. #<<
#' @param arg2 a string to be pasted; default is NA. #<<
#'
#' @return
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---


### `@return` describes what the function returns

```{r, eval=F,highlight=T}
#' my_function() 
#'
#' This function pastes together two inputs in a string 
#'
#' @param arg1 a string to be pasted; default is NA. 
#' @param arg2 a string to be pasted; default is NA. 
#'
#' @return a string #<<
#' @export
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---


### `@export` indicates that the function should be exported into the package namespace (i.e. the function can be called when the package is imported)

```{r, eval=F,highlight=T}
#' my_function() 
#'
#' This function pastes together two inputs in a string 
#'
#' @param arg1 a string to be pasted; default is NA. 
#' @param arg2 a string to be pasted; default is NA. 
#'
#' @return a string 
#' @export #<<
#'
#' @examples
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---


### `@examples` offers an example of how the function should be executed. You should be able to copy and paste this example into the R console and it should run. 

```{r, eval=F,highlight=T}
#' my_function() 
#'
#' This function pastes together two inputs in a string 
#'
#' @param arg1 a string to be pasted; default is NA. 
#' @param arg2 a string to be pasted; default is NA. 
#'
#' @return a string 
#' @export 
#'
#' @examples #<<
#' #<<
#' my_function("hello","cat") #<<
#' #<<
my_function <- function(arg1=NA,arg2=NA){
  print(paste(arg1,arg2,sep="_")) 
}
```

---

class: inverse, center, middle

# Testing

---

### Testing is a vital part of package development. It ensures that your code does what you want it to do. 

--

### Why design tests?

- Fewer bugs
- More robust code
- Peace of mind as you improve/alter functionality

--

### How it works?

- `testthat` makes a special folder in your package directory. 
- Tests are composed of **expectations** organized into individual files.
- The output from your code is tested against the expected output. If it matches, it passes; otherwise, it throws an error. 

---

## Setting up a test

### (1) Context

> Group together related tests with a context. E.g. "these tests all deal with how data is ingested"

```r
context("Data prepping functions")
```

---

## Setting up a test

### (1) Context

### (2) Expectation

> What output should one expect your function to produce? 

```r
a <- 10
testthat::expect_equal(a, 10)
```

```r
testthat::expect_identical(sqrt(2) ^ 2, 2)
```

---

## Setting up a test

### (1) Context

### (2) Expectation

### (3) Test

> group together multiple expecations to test the output from a function

```r
context("String length")
library(stringr)

test_that("str_length is number of characters", {
  expect_equal(str_length("a"), 1)
  expect_equal(str_length("ab"), 2)
  expect_equal(str_length("abc"), 3)
})
```

[_example source_](https://r-pkgs.org/tests.html)

---

class: inverse, center, middle

# Distribution

---

## Sharing Your Package with the World

### Two (most popular) ways of doing this...

--

### (1) Github

> Upload your package repository to Github. Anyone can install from a public repo. 
```r
devtools::install_github('your_account/your_package')
```

> Use Github features like readme and pages to create clean, detailed vignettes and guidance on how to user your package. 

--

### (2) Comprehensive R Archive Network (CRAN)

> Build your source package `.tar.gz` and submit via the CRAN portal (https://cran.r-project.org/submit.html)

---

## When to submit your package to CRAN

### It's stable

> Code is (relatively) bug free and runs without issues on different machines (mac/windows/linux)

--

### It's generalized

> Package operates across an array of problem domains and object types (i.e. you've built logic for dealing with different class types). 

--

### It's distinct

> Package offers novel functionality not already contained within an existing package. If not, then your package is more personalized and should just live on github.

---

## When to submit your package to CRAN



### Remember...

#### CRAN is a marketplace &rarr; Having a package on CRAN doesn't mean anything if no one uses it. 
#### Follow trends and conventions. 
#### Think about the user. 
#### Design code that plays along with existing (widely adopted) workflows (e.g. tidy ecosystem). 

---

# Example &rarr; [tidysynth](https://github.com/edunford/tidysynth)

.center[
### [https://github.com/edunford/tidysynth](https://github.com/edunford/tidysynth)
]

<br>

.center[
<img src = "figures/tidysynth_logo.png" width=40%, height=40%>
]

---

<br>
.center[
# Thanks!
<br>

.center[
### Eric Dunford 
_McCourt School of Public Policy_<br>
**_Georgetown University_**<br>
eric.dunford@georgetown.edu<br>
[ericdunford.com](http://ericdunford.com/)
]
]

